sdmTMB: an R package for fast, flexible, and user-friendly generalized linear mixed effects models with spatial and spatiotemporal random fields

### 2022-11-15 doc/paper-jss/paper-jss.Rmd
Edits

Data are often collected in space or in space repeatedly over time.
While such data are a rich source of information [e.g., @legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself by, for example, identifying locations with shared responses [e.g., @thorson2019d; @barnett2021].

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2022].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}\-pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].

>>>>

Data are often collected in space or in space repeatedly over time.
While such data are a rich source of information [e.g., @legendre1989a; @rossi1992; @tilman1997spatial], they are challenging to properly model---data closer in space and time are usually more similar to each other than data farther apart due to measured and unmeasured variables [@cressie1993; @diggle2007; @cressie2011].
While measured variables can be accounted for with predictors in a model (e.g., measuring and modelling temperature effects on species abundance), unmeasured variables (e.g., everything influencing species abundance but not explicitly modelled) can cause residual spatial correlation.
Accounting for this residual correlation is important because doing so allows for valid statistical inference [@legendre1989a; @dormann2007], can improve predictions  [e.g., @shelton2014], and can be of interest itself [e.g., @thorson2019d; @barnett2021].

@@

Such models quickly become computationally limiting due to the need to invert large matrices to keep track of covariation among data [e.g., @rue2005gmrf; @latimer2009].
Many solutions have been proposed, such as predictive processes [@banerjee2008; @latimer2009], the stochastic partial differential equation (SPDE) approximation to GRFs [@lindgren2011], and nearest-neighbour Gaussian processes [@datta2016; @finley2022].
These approaches reduce the scale of the covariance estimation problem while providing a means to evaluate the data likelihood, thereby allowing fitting via Bayesian [@gelfand2017] or maximum likelihood methods.
This can greatly improve computational efficiency [e.g., @heaton2019].
The SPDE approach is a solution popularized via the \proglang{R}-\pkg{INLA} \proglang{R} package [@rue2009; @lindgren2011; @lindgren2015] and an implementation in \pkg{TMB} [Template Model Builder, @kristensen2016] that partially relies on \proglang{R}-\pkg{INLA} to create input matrices [e.g., @thorson2019; @osgood-zimmerman2021].
Details are beyond the scope of this paper and are not necessary to use the software discussed here, but the idea is that the solution to a specific SPDE is a GRF with a Matérn covariance function and this 'trick' enables one to efficiently fit approximations to GRFs to large spatial datasets [@lindgren2011].
